---
title: "Assignment 09: Data Scraping"
author: "Ariel O'Callaghan"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(lubridate)

#install.packages("rvest")
library(rvest)

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy=TRUE)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

pwsid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid
  
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td , th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
# Order of months from scraped data. Month <- c(1,5,9,2,6,10,3,7,11,4,8,12)

df_withdrawals<-data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                          "Year"= rep(2021,12),
                          "Max_Withdrawls_mgd"= as.numeric(max.withdrawals.mgd))

df_withdrawals <- df_withdrawals %>% 
  mutate(Water_system_name = !!water.system.name,
         PWSID = !!pwsid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))
class(Date)
#5
#plot geom_line(aes(group=1))
Durham.max.withdrawl.plot<-ggplot(df_withdrawals) +
  geom_line(aes(x= Date,y= as.numeric(max.withdrawals.mgd))) + 
  labs(title = "2021 Water usage data for",
       subtitle = water.system.name,
       y="Withdrawal (mgd)",
       x="Date")
print(Durham.max.withdrawl.plot)

  
 
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.

scrape.it <- function(the_year, the_pwsid ) {
#https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021'

the_website <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", the_pwsid,'&year=', the_year ))
           
water.system.name <- the_website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

pwsid <- the_website %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid

ownership <- the_website %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.withdrawals.mgd <- the_website %>% 
  html_nodes("th~ td+ td , th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

#rep(1:12)
df_scrape<-data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                          "Year"= rep(the_year,12),
                          "Max_Withdrawls_mgd"= as.numeric(max.withdrawals.mgd))%>% 
  mutate(Water_system_name = !!water.system.name,
         PWSID = !!pwsid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))

return(df_scrape)

}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

df_durham_2015 <- scrape.it(2015, '03-32-010')
view(df_durham_2015)

df_durham_2015$Date <- as.Date(df_durham_2015$Date, format ="%Y-%M-%D")
class(df_durham_2015$Date)

class(df_durham_2015$Max_Withdrawls_mgd)
class(df_durham_2015$Date)

Durham_2015_plot<-ggplot(df_durham_2015,aes(x=Date,y=Max_Withdrawls_mgd, group=1)) +
  geom_line()+
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Max Daily Water Withdrawal Durham"), X = "Date", y = " Max Daily Withdrawals (mgd)")
print(Durham_2015_plot)
#geom_smooth(method="loess",se=FALSE)
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8

df_Asheville_2015 <- scrape.it(2015,'01-11-010')
view(df_Asheville_2015)

durham_asheville<- rbind(df_Asheville_2015, df_durham_2015)


durham_asheville$Date <- as.Date(durham_asheville$Date, format ="%Y-%M-%D")
class(durham_asheville$Date)
class(durham_asheville$Max_Withdrawls_mgd)


Durham__ashville_plot<-ggplot(durham_asheville,)+
  geom_line(aes(x=Date,y=Max_Withdrawls_mgd, color=Water_system_name))
  labs(title = paste("2015 Max Daily Water Withdrawal Durham and Asheville"), X = "Date", y = " Max Daily Withdrawals (mgd)")
print(Durham__ashville_plot)



```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9
Ash_pwsid <- "01-11-010"
Ash_year <- rep(2010:2019)
Asheville_9y_df <- map2(Ash_year, Ash_pwsid, scrape.it)
Ash_bind <- bind_rows(Asheville_9y_df)

Ash_bind$Date <- as.Date(Ash_bind$Date, format ="%Y-%M-%D")
class(Ash_bind$Date)

Ashvile_Durham_plot_9 <- ggplot(Ash_bind, aes(x=Date,y=Max_Withdrawls_mgd))+
geom_line()+
labs(title = paste("Asheville's max daily withdrawal 2010-2019"), X = "Date", y = " Max Withdrawls (MGM)")
print(Ashvile_Durham_plot_9)



```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
Asheville uses more water over time. There is a peak water usage around 2017. There seems to be many seasonal and yearly fluxuuations but the max waer usage started to increase around 2017.  
