---
title: 'Assignment 3: Data Exploration'
author: "Ariel O'Callaghan"
output:
  html_document:
    df_print: paged
  pdf_document: default
geometry: margin=2.54cm
---

## OVERVIEW

#This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

The completed exercise is due on Sept 30th.

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}

#setwd("/home/guest/R/EDA-Fall2022) 
#change wd from assingments to main eda folder

#install.packages("tinytex")
#tinytex::install_tinytex()

#Allow for code and comments to not run off page when knitting file. Note- add to top of every rmd

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)


#installing necessary packages, one installed add # to knit assignment 
#install.packages("tidyverse")
#install.packages("dplyr")
library(ggplot2)
#install.packages("lubridate")
library(lubridate)
#adding datasets needed for assingment

Neonics.data <-read.csv("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter.data <-read.csv("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

#str(Neonics.data$Effect)

#Explore dataset
#View(Neonics.data) 
#class(Neonics.data) #data.frame
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: This database provides information on adverse effects of single chemical stressors aquadic and terrestiral species. Inspect are important for polluation and food production and understanding stressors are important. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Understanding forest litter and woody debris helps us better understand the health of our mountains. This monitoirng allows us to understand diverse ecosystems in multiplal spatial and temporral scales. It also helps better understand the biodiversity in our ecosystem. Litter and woody debris sampling help provides understanding about plans. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter and fine woody debris are collected from elevated ground traps
 2. Sampling desing ecompases spaital paramaters such as NEON sites with woody vegitation >2M. 
 3. Ground traps are sampled once per year and more frequenly depending on density of vegitation and forest. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
colnames(Neonics.data)
na.omit(Neonics.data)
na.omit(Litter.data)
str(Neonics.data)
dim(Neonics.data)
length(Neonics.data)

#saving new datasets
#write.csv(Neonics.data, file = "./Data/Ariel_processed/Neonics.data.v2", row.names=FALSE)
#write.csv(Litter.data, file = "./Data/Ariel_processed/Litter.data.v2", row.names=FALSE)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics.data$Effect)
summary(Neonics.data$Effect.Measurement)
```

> Answer: The main effects in this dataset are mortality, feeding behavior, population, behavior and mortality. These are all important aspects to study for species survial and better understanding population estimates. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics.data$Species.Common.Name)
```

> Answer:
The most common are Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. Most of the top insects sudied are bees which are pollnators and a keystone species for our food production. 

8.  Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics.data$Conc.1..Author.)

```

> Answer:The  class(Neonics.data$Conc.1..Author.) shows up as "factor" because it is not a numerical data set with continuous or descrete values. This column is a categorical dataset. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics.data)+
  geom_freqpoly(aes(x = Publication.Year), bins=15)
ggplot(Neonics.data)+
  geom_freqpoly(aes(x = Publication.Year), bins=25)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics.data)+
  geom_freqpoly(aes(x = Publication.Year), bins=15, color = "red")

ggplot(Neonics.data)+
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins=15)


```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common testing locations are the labs and natural field. The lab testing only supparsed natral field around 2009 and beacame the most common test location. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information.

```{r}
ggplot(Neonics.data, aes(x=Endpoint)) +
  geom_bar() 
  
  
#label_parsed(Endpoint, multi_line = TRUE) did not work try to space text on X axis

```

> Answer:The two most common end points are LOEC and NEOL. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
#help("as.Date")
#Litter.data <-read.csv("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
class(Litter.data$collectDate) # checking current class
Litter.data$collectDate <- c(ymd(Litter.data$collectDate)) #Changing to date format
class(Litter.data$collectDate) #checking the changed class
unique(Litter.data$collectDate)

#Litter.data.datetime <- as.Date.factor(Litter.data$collectDate, format= "%Y/%m/%d") - Do not use
#na.omit(Litter.data.datetime)
#format(Litter.data$collectDate, format= "%Y/%m/%d") do not use
#class(Litter.data.datetime$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
#help("unique")
#help(summary)
unique(Litter.data$plotID)
```

> Answer:The uniuqe function returns a vector, data frame or array with dupliance elements/rows removed. The summary is a generic function which produces summaries of the various functions. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

ggplot(Litter.data, aes(x =functionalGroup)) +
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter.data) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))

ggplot(Litter.data) +
  geom_violin(aes(x = dryMass, y = functionalGroup), draw_quantiles = c(0.25, 0.5, 0.75)) 

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplot allows you to see the were the majority of the groupings are for the functional groups by drymass but also the outliners. The violin plots which display density distribution are not as effective. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles have the highest biopass at these sights. 
